Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-09-22 article.Rmd
Change title to prediction
Typos and minor edits

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection"
author: "Cameron Blevins and Lincoln Mullen"
output: html_document
---

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the AHR publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the AHR makes it an ideal source to analyze the historical profession on a large scale.

We take the AHR as a rough proxy for gatekeeping within the larger historical profession. When a book appears in the AHR it serves as a signal that other historians in positions of power have "approved" it for consideration. This signal is not necessarily correlated with the quality of the book. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review. On the flip side, a book that does not appear in the AHR may have gone on to have a significant impact on the profession. It is precisely this professional gatekeeping dimension that makes the AHR useful to study in the context of gender representation.

Scraping the table-of-contents of every issue of the American Historical Review results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could then use to plot the proportion of female and male authors over time. The temporal trajectory of gender representations roughly resembles that of history dissertations: women began making inroads in the late 1960s and have made steady gains over the past four decades. By the twenty-first century the proportion of female authors reviewed in the AHR had more than tripled.

@@

But a closer look shows important differences between the proportion of women writing history dissertations and the proportion of women appearing in the AHR. Although both have trended upwards since the 1970s the slope is a lot steeper for dissertations than it is for the AHR. Female historians have very nearly closed the gaps in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of what happens from that point onwards. In book reviews published in the AHR male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over in the pages of the historical profession's leading journal.

@@

The path from a dissertation to an AHR book review is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the AHR editors for review. As of August 2014 the journal’s website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014]. Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More insidiously, women continue to face insidious gender discrimination.

We then used our program to infer the gender of not just book authors reviewed by the AHR, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the AHR.

@@

The American Historical Association also categorizes their reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the AHR: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

>>>>

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction"
author: "Cameron Blevins and Lincoln Mullen"
output: html_document
---

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the *AHR* publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." The *AHR* is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the *AHR* makes it an ideal source to analyze the historical profession on a large scale.

We take the <i>American Historical Review</i> as a rough proxy for gatekeeping within the larger historical profession. When a book appears in the *AHR* it serves as a signal that other historians in positions of power have "approved" it for consideration. This signal is not necessarily correlated with the quality of the book. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review. On the flip side, a book that does not appear in the *AHR* may have gone on to have a significant impact on the profession. It is precisely this professional gatekeeping dimension that makes the AHR useful to study in the context of gender representation.

Scraping the table-of-contents of every issue of the American Historical Review results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could then use to plot the proportion of female and male authors over time. The temporal trajectory of gender representations roughly resembles that of history dissertations: women began making inroads in the late 1960s and have made steady gains over the past four decades. By the twenty-first century the proportion of female authors reviewed in the *AHR* had more than tripled.

@@

But a closer look shows important differences between the proportion of women writing history dissertations and the proportion of women appearing in the *AHR*. Although both have trended upwards since the 1970s the slope is a lot steeper for dissertations than it is for the *AHR*. Female historians have very nearly closed the gap in terms of newly completed dissertations, but the glass ceiling remains stubbornly low in terms of what happens from that point onwards. In book reviews published in the *AHR* male authors continue to outnumber female authors by a factor of nearly 2 to 1. The long-term gains made by female doctoral students do not carry over in the pages of the historical profession's leading journal.

@@

The path from a dissertation to a book review in the *American Historical Association* is not an automatic one. Not all dissertations are published as books, not all books are submitted to the American Historical Review, and not all book submissions are accepted by the *AHR* editors for review. As of August 2014 the journal’s website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)."^[http://www.indiana.edu/~ahrweb/book_review_guide.html - retrieved August 25 2014]. Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face insidious gender discrimination.

We then used our program to infer the gender of not just book authors reviewed by the *AHR*, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

The American Historical Association also categorizes their reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the *AHR*: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.
